Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-10-14 paper/main.tex

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if for each set of
  states $S\subseteq Q$ it holds that
  \begin{enumerate}[A.]
  \item\label{en:1} all transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$) or
  \item\label{en:2} all transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$) or
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transtition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transtion leading from a state in $S$
    to a state outside $S$ and a transtion leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transtitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not holds for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradition that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
/  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.\qed
\end{proof}

@@

The second optimization is specific for $\Phi_{3,C,K}$. In
the construction of $\mcycle_3(\vec{e})$, we replace the subformula
$\bigvee_{t\in\delta}e_t$ enforcing the nonemptiness of $\Te$ by
$\mcycle_2(\vec{e})$. This modification does not change the overall
semantics of $\mcycle_3(\vec{e})$, but our prelimnary experiments
showed that QBF solvers can often solve the modified formula
$\Phi_{3,C,K}$ faster.

>>>>

\subsection{Strict version $\mcycle_3(\vec{e})$}
Before we give the definition of $\mcycle_3(\vec{e})$, we prove that
cycles can be characterised in the following way.
\begin{lemma}
  A nonempty set $T\subseteq\delta$ is a cycle if and only if for each set of
  states $S\subseteq Q$ it holds that
  \begin{enumerate}[A.]
  \item\label{en:1} all transitions in $T$ lead from a state in $S$ to a state in $S$ (i.e., $T\subseteq\delta_S$) or
  \item\label{en:2} all transitions in $T$ lead from a state outside $S$ to a state outside $S$ (i.e., $T\subseteq\delta_{Q\smallsetminus S}$) or
  \item\label{en:3} $T$ contains a transition leading from a state in $S$ to a state outside $S$ and a transition leading from a state outside
    $S$ to a state in $S$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  We first prove the direction ``$\Longrightarrow$''. Let $T$ be a
  cycle and $S\subseteq Q$ be an arbitrary set of states. We show that
  if (\ref{en:1}) and (\ref{en:2}) do not hold, then (\ref{en:3}) has
  to hold. Hence, assume that $T\not\subseteq\delta_S$ and
  $T\not\subseteq\delta_{Q\smallsetminus S}$. Then there are two
  cases.
  \begin{itemize}
  \item $T$ contains a transition $t\in\delta_S$ and a transition
    $t'\in\delta_{Q\smallsetminus S}$. The definition of a cycle
    implies that there exists a path $t_1t_2\ldots t_n\in T^+$ from a
    state $p$ back to $p$ containing both $t$ and $t'$. However, this
    implies that $T$ contains a transition leading from a state in $S$
    to a state outside $S$ and a transition leading from a state outside $S$
    to a state in $S$.
  \item $T$ contains a transition $t$ leading from a state in $S$ to a
    state outside $S$ (or vice versa). However, as $T$ is a cycle,
    there exists a path $t_1t_2\ldots t_n\in T^+$ that leads from a
    state $p$ to the same state and contains $t$. Hence, $T$ has to
    contain also a transition leading from a state outside $S$ to a
    state in $S$ (or vice versa).
  \end{itemize}
  In both cases, (\ref{en:3}) holds.

  Now we prove the opposite direction ``$\Longleftarrow$'' by
  contraposition. Assume that a nonempty set $T$ is not a cycle. We
  show that there is a set $S\subseteq Q$ such that neither
  (\ref{en:1}) nor (\ref{en:2}) nor (\ref{en:3}) holds. Let $p$ be a
  state such that some transition of $T$ leads from $p$. We define the
  set $\post$ of states reachable from $p$ via transitions in $T$ and the set
  $\pre$ of states from which $p$ is reachable via transitions of $T$.
  \[
    \begin{array}{r}
    \post=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }p\textrm{ to }q\}\\[.5ex]
    \pre=\{p\}\cup\{q\in Q\mid \textrm{there is a path in }T^+\textrm{ from }q\textrm{ to }p\}
    \end{array}
  \]  
  As there is a transition of $T$ leading from $p$, we have that
  $T\not\subseteq\delta_{Q\smallsetminus\post}$ and
  $T\not\subseteq\delta_{Q\smallsetminus\pre}$, i.e., (\ref{en:2})
  does not holds for $\post$ and $\pre$. Further, the definition of
  $\post$ implies that there is no transition of $T$ leading from a
  state in $\post$ to a state outside $\post$, which means that
  (\ref{en:3}) does not hold for $\post$. Similarly, $T$ contains no
  transition leading from a state outside $\pre$ to a state in $\pre$,
  which means that (\ref{en:3}) does not hold for $\pre$. Now we prove
  by contradition that (\ref{en:1}) does not hold for at least one of
  $\post,\pre$. Hence, let us assume that $T\subseteq\delta_\post$ and
  $T\subseteq\delta_\pre$. Then for each $t_i\in T$ leading from $p_i$
  to $q_i$ we have that $p_i\in\post$ and $q_i\in\pre$, which implies
  that
  \begin{itemize}
  \item $p_i=p$ (we set $\rho'_i=\varepsilon$ in this case) or
    there is a path $\rho'_i\in T^+$ leading from $p$ to $p_i$, and
  \item $q_i=p$ (we set $\rho''_i=\varepsilon$ in this case) or
    there is a path $\rho''_i\in T^+$ leading from $q_i$ to $p$.
  \end{itemize}
  Then there is a path $\rho_i=\rho'_it_i\rho''_i\in T^+$ leading from
  $p$ back to $p$ and containing $t_i$. If we concatenate all these
/  paths, we get the path $\rho_1\rho_2\ldots\rho_{|T|}\in T^+$ that
  contains all transitions of $T$ and leads from $p$ back to $p$,
  which means that $T$ is a cycle.  This is a contradiction.\qed
\end{proof}

@@

The second optimization is specific for $\Phi_{3,C,K}$. In the
construction of $\mcycle_3(\vec{e})$, we replace the subformula
$\bigvee_{t\in\delta}e_t$ enforcing the nonemptiness of $\Te$ by
$\mcycle_2(\vec{e})$. This modification does not change the overall
semantics of $\mcycle_3(\vec{e})$, but our preliminary experiments
showed that QBF solvers can often solve the modified formula
$\Phi_{3,C,K}$ faster.
